---
title: "Assignment1"
author: "Mireille Vargas"
date: "'r format(Sys.Date(), '%B %d, %Y')'"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/mirei/OneDrive/Documents/GitHub/ESS218X")
```


```{r}
library(tidyverse)
library(plotly)
library(sf)
library(tigris)
library(leaflet)
library("dplyr")
library("plyr")
library("readr")
library(censusapi)
library("rgdal")
library("sp")
library(zoo)
```

## Including Plots

You can also embed plots, for example:

```{r}
years <- 2017:2020
quarters <- 1:4
types <- c("Electric", "Gas")

pge_Elec_Gas_All <- NULL

for (quarter in quarters){
  for (year in years){
    #if year == 2020 #if statement to seperate 2020 quarters 1 and 2, would say if 2020 and quarter is 3 or 4 then next
    if (year == 2020 & quarter == 3){
      next()
    }
    if (year == 2020 & quarter == 4){
      next()
    }
    for (type in types){
      filename <-
        paste0(
          "PGE_",
          year,
          "_Q",
          quarter,
          "_",
          type,
          "UsageByZip.csv"
        )
      #print(filename)
    temp <- read_csv(filename)
temp$TOTALkBTU <- NULL
temp$AVERAGEkBTU <- NULL
temp$DATE <- as.yearmon(paste(temp$YEAR, temp$MONTH), "%Y %m")
if (type == "Gas"){
  temp$TOTALkBTU <- temp$TOTALTHM * 100 
  temp$AVERAGEkBTU <- temp$TOTALkBTU/temp$TOTALCUSTOMERS
  temp$TOTALTHM <- NULL
  temp$AVERAGETHM <- NULL
}
if (type == "Electric"){
  temp$TOTALkBTU <- temp$TOTALKWH * 3.4121416
  temp$AVERAGEkBTU <- temp$TOTALkBTU/temp$TOTALCUSTOMERS
  temp$TOTALKWH <- NULL
  temp$AVERAGEKWH <- NULL
}
pge_Elec_Gas_All <- rbind(pge_Elec_Gas_All, temp)
    
    }
  }
saveRDS(pge_Elec_Gas_All, "pge_Elec_Gas_All.rds")  
}
head("pge_Elec_Gas_All.rds")
```

Filtering the data to have only the Bay Area zipcodes. 
```{r}
options(tigris_use_cache = FALSE)
ca_counties <- counties("CA", cb = T, progress_bar = F)
#st_drivers()
bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  ca_counties %>%
  filter(NAME %in% bay_county_names)

usa_zips <- 
  zctas(cb = T, progress_bar = F)
#creates a dataframe of all bay area zipcodes
bay_zips <-
  usa_zips %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_set_geometry(NULL) %>% 
  left_join(usa_zips %>% select(GEOID10)) %>% 
  st_as_sf()

# filter(pge_Elec_Gas_All$ZIPCODE %in% bay_zips$ZCTA5CE10)
bayAreaPGE<- subset(pge_Elec_Gas_All, pge_Elec_Gas_All$ZIPCODE %in% bay_zips$ZCTA5CE10)
```

Creating the filtered data set.
```{r}
bayAreaPGE_elec_final <-
  bayAreaPGE %>%
  # subset(pge_Elec_Gas_All, pge_Elec_Gas_All$ZIPCODE %in% bay_zips$ZCTA5CE10)%>%
  filter( #choose only these customer classes
    CUSTOMERCLASS %in%
      c(
        "Elec- Residential",
        "Elec- Commercial", "Gas- Residential", "Gas- Commercial" #I could add gas data in here but I would have to change the subset
      )

  ) %>%
  select(
    #c(DATE, CUSTOMERCLASS, TOTALCUSTOMERS, TOTALkBTU) #include these columns
    # !c(YEAR, MONTH, COMBINED, AVERAGEkBTU)
    !c(COMBINED, AVERAGEkBTU)
  ) %>%
  #group_by(DATE, CUSTOMERCLASS) %>%
  dplyr::group_by(DATE, CUSTOMERCLASS) %>%
  #summarize(
  dplyr::summarise(
    TOTALkBTU =
      sum(
        TOTALkBTU,
        na.rm = T
      ),
    TOTALCUSTOMERS =
      sum(
        TOTALCUSTOMERS,
        na.rm = T
      )
  ) %>%
  mutate(
    AVERAGEkBTU =
      TOTALkBTU/TOTALCUSTOMERS
  )

bayAreaPGE_elec_final

```
Create the graph
```{r}
pge_chart <-
  bayAreaPGE_elec_final %>% 
  ggplot() +
  geom_bar(
    aes(
      x = DATE %>% factor(),
      y = TOTALkBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Date",
    y = "kBTU",
    title = "PG&E Bay Area Monthly Electricity and Gas Usage, 2017-2020",
    fill = "Customer Class Type"
  ) + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)#to shift the dates so they're visible
    )

pge_chart

pge_chart %>% ggplotly()

```

Creating the dataset of COVID influence 
one of the average KBTU consumption per month before covid (Mar-Jun), and one of the 2020 KBTU consumption in Mar-Jun. is there a way to combine the two data frames by subtracting the covid-era consumption values from pre-covid consumption (do left_join)
```{r}
#group by zipcode
#percent change: (2020-2019/2019 * 100)
bayAreaCOVID <-
  bayAreaPGE %>%
  filter(
    YEAR %in% (2019:2020), MONTH %in% (1:5), CUSTOMERCLASS == "Elec- Residential"
    ) %>%
  mutate(
    ZIPCODE = ZIPCODE %>% as.character()
    ) %>%
  dplyr::group_by(ZIPCODE, YEAR) %>%
  dplyr::summarise(
    TOTALkBTU =
      sum(
        TOTALkBTU,
        na.rm = T
      )
  ) %>%
  right_join(
    bay_zips %>% select(GEOID10),
    by = c("ZIPCODE" = "GEOID10")
  ) %>%
  pivot_wider(
    names_from = YEAR,
    names_prefix = "kBTU",
    values_from = TOTALkBTU
  ) %>%
  dplyr::filter(kBTU2019> 0) %>%
  mutate(
    COVID_influence =
      ((kBTU2019 - kBTU2020)/ kBTU2019) * 100
  ) %>%
  # drop_na()%>%
  st_as_sf()%>%
  st_transform(4326) #%>%
  # st_set_geometry(NULL) 


bayAreaCOVID
```

Creating the map
```{r}
# class(bayAreaCOVID)
res_pal <- colorNumeric(
  palette = "Blues",
  domain =
    bayAreaCOVID$COVID_influence
)

leaflet() %>% #leaflet wants a spatial object
  addTiles() %>%
  addPolygons(
    data = bayAreaCOVID,
    fillColor = ~res_pal(COVID_influence),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(COVID_influence),
      "kBTU Percent Change in",
      ZIPCODE
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  )%>%
  addLegend(
    data = bayAreaCOVID,
    pal = res_pal,
    values = ~COVID_influence,
    title = "Percent Change in Residential<br>Electric Energy Usage possibly<br>due to COVID-19"
)
```

